---
title: "Assignment2_Jo"
author: "Josue Iradukunda"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Import the CSV (comma-separated value) using the `read.csv()` command. Name it `survey_data`.
```{r echo=FALSE}
survey_data <- read.csv("data.csv")
View(survey_data)
```

# Data cleaning using grepl

```{r echo=FALSE}
static <- grepl("statistic", survey_data$Background, ignore.case = TRUE)
survey_data$Background[static] <- "Applied Statistics"
```

```{r echo=FALSE}
data_sc <- grepl("Data Scientist", survey_data$Background, ignore.case = TRUE)
survey_data$Background[data_sc] <- "Data Science"
```

```{r echo=FALSE}
ind_art_desgn <- grepl("Industry art and design","Design", survey_data$Background, ignore.case = TRUE)
survey_data$Background[ind_art_desgn] <- "Industrial Art and Design"
```

```{r echo=FALSE}
ind_art_desgn <- grepl("Design", survey_data$Background, ignore.case = TRUE)
survey_data$Background[ind_art_desgn] <- "Industrial Art and Design"
```

```{r echo=FALSE}
infor_sys <- grepl("information syst", survey_data$Background, ignore.case = TRUE)
survey_data$Background[infor_sys] <- "Information System"
```

```{r echo=FALSE}
civ_eng <- grepl("civil engin", survey_data$Background, ignore.case = TRUE)
survey_data$Background[civ_eng] <- "Civil Engineering"
```

```{r echo=FALSE}
quant_surv <- grepl("Quantity surv", survey_data$Background, ignore.case = TRUE)
survey_data$Background[quant_surv] <- "Quantity Surveying"
```

```{r echo=FALSE}
trav_tour <- grepl("I have complete my bachelor degree in travel amd and tourism management", survey_data$Background, ignore.case = TRUE)
survey_data$Background[trav_tour] <- "Travel and Tourism Management"
```

```{r echo=FALSE}
demog <- grepl("demogr", survey_data$Background, ignore.case = TRUE)
survey_data$Background[demog] <- "Demography"
```

```{r echo=FALSE}
IT <- grepl("Information Technology Major of Information Management", survey_data$Background, ignore.case = TRUE)
survey_data$Background[IT] <- "Information Technology"
```

```{r echo=FALSE}
travel <- grepl("Bachelor degree in Travel and Tourism management at UTB.", survey_data$Background, ignore.case = TRUE)
survey_data$Background[travel] <- "Travel and Tourism Management"
```

```{r echo=FALSE}
travel <- grepl("Tourism", survey_data$Background, ignore.case = TRUE)
survey_data$Background[travel] <- "Travel and Tourism Management"
```

```{r echo=FALSE}
dt_table <- table(survey_data$Background)
data_summary <- as.data.frame(dt_table)
colnames(data_summary) <- c("Field","Count")

data_summary$Percentage <- round(data_summary$Count/sum(data_summary$Count) * 100, 2)
data_summary <- data_summary[order(-data_summary$Count),]
data_summary
```

## 
The table illustrates the distribution of respondents across nine fields of study, with **Applied Statistics** leading with 27 respondents, accounting for 52.94% of the total. **Data Science** comes next with 5 respondents (9.80%), while both **Industrial Art and Design** and **Information System** each have 4 respondents (7.84%). Fields like **Civil Engineering**, **Quantity Surveying**, and **Travel and Tourism Management** each have 3 respondents (5.88%). The least represented fields, **Demography** and **Information Technology**, each have only 1 respondent, making up 1.96%. In summary, the data is predominantly skewed toward Applied Statistics, with most other fields showing limited participation.

```{r echo=FALSE}
pie(data_summary$Count, labels = paste(data_summary$Field," (",data_summary$Percentage, "%)", sep = ""), main = "Distrution of Background", cex= 0.8, col = rainbow(length(data_summary$Field)))
```
