---
title: "Assignement 2"
output:
  word_document: default
  pdf_document: default
date: "2024-01-10"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Read the dataset in R.

```{r echo=FALSE}
# required package
library(haven)

# import data
data_sav <- read_sav("C:\\Users\\PMVANO\\Desktop\\GROUP ASSIGNMENT\\DHS2020\\RWPR81FL.SAV")
#data_map <- read.table("C:\\Users\\PMVANO\\Desktop\\GROUP ASSIGNMENT\\DHS2020\\RWPR81FL.SAV", sep = "\t", header = TRUE)

# data dimension
data_sav
```

 Getting familiar with the data
 
```{r echo=FALSE}
# your code here
View(data_sav)
dim_no <- dim(data_sav)
#In our dataset `r dim_no[1]` is total number of Obersvation and `r dim_no[2]` total number of Variables.
 #View(data_map)

```

2. Selecting few columns, 

```{r echo=FALSE}
selected_columns <- data_sav[, c("HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7","HML10", "HML22", "HML32","HML33", "HML35")]
selected_columns

```
3. Rename variables using the variable descriptions for  meaningful (short) name

```{r echo=FALSE}
#Rename variables using the variable descriptions given.
colnames(selected_columns) <- c(
  "ClusterNumber", "HouseholdMembers", "EligibleWomen", "EligibleMen", "ChildrenUnder5", 
  "District", "Region", "ResidenceType", "Altitude", "BedNetPresence", 
  "ChildrenUnder5BedNet", "WealthIndex", "Age", "EducationLevel", 
  "NetObserved", "NetObtainedMonthsAgo", "NetBrand", "ITN", "NetSource", 
  "BloodSmearMalariaResult", "MalariaMeasurementResult", "RapidTestMalariaResult"
)
head(selected_columns)

```


## Data cleaning

1. Inspecting each variables,then  decoding variable to its original unique variables. 

```{r echo=FALSE}
# Decode all columns
all_columns <-  c(
  "ClusterNumber", "HouseholdMembers", "EligibleWomen", "EligibleMen", "ChildrenUnder5", "District", "Region", "ResidenceType", "Altitude", "BedNetPresence","ChildrenUnder5BedNet", "WealthIndex", "Age", "EducationLevel", 
  "NetObserved", "NetObtainedMonthsAgo", "NetBrand", "ITN", "NetSource", 
  "MalariaMeasurementResult", "BloodSmearMalariaResult", "RapidTestMalariaResult")

selected_columns[all_columns] <- sapply(selected_columns[all_columns], as_factor)
selected_columns
```

2.HANDLING MISSING VALUES

```{r echo=FALSE}
# checking columns with missing values in entire datasets
missing_value <- colSums(is.na(selected_columns))
# Determining columns with missing values
missing_value[missing_value > 0]
```
REPLACING MISSING VALUES
```{r echo=FALSE}
# replacing missing values with mode for catergorical variables  
# Mode function to handle categorical data
calculate_mode <- function(x) {
  # Handle NA values
  x <- x[!is.na(x)]
  # Find the most frequent value
  if (length(x) == 0) return(NA)  # If all values are NA, return NA
  mode_value <- names(sort(table(x), decreasing = TRUE))[1]
  return(mode_value)
}
# Replacing mean for numeric variables,Handle all using lappy function
selected_columns[] <- lapply(selected_columns, function(col) {
  if (is.numeric(col)) {
    # Replace numerical missing values with the mean
    col[is.na(col)] <- mean(col, na.rm = TRUE)
  } else if (is.factor(col) || is.character(col)) {
    # Replace categorical missing values with the mode
    col[is.na(col)] <- calculate_mode(col)
  }
  return(col)
})

selected_columns
```

3. Create new variables

```{r echo=FALSE}
#a. Creating new variable called "Old Mosquito"binary with 1 when mosquito is more than 24 months old. 
selected_columns<- selected_columns %>%
  mutate(Old_Mosquito = ifelse(NetObtainedMonthsAgo > 24, 1, 0))

# View the updated data
selected_columns
```

 "Average District altitude".

```{r echo=FALSE}
#b.Creating "Average District altitude" Variable

selected_columns<- selected_columns %>%
  group_by(District) %>%
  mutate(Average_District_Altitude = mean(as.numeric(Altitude))) %>% ungroup()

# View the data
print(selected_columns)
```
 Data visualizations
1.Bar plot
```{r}
bed_net_by_region <- selected_columns %>%
  group_by(Region, BedNetPresence) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  ggplot(aes(x = as.character(Region), y = Count, fill = as.character(BedNetPresence))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Bed Net Presence by Region",
       x = "Region", 
       y = "Number of Households",
       fill = "Bed Net Presence") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(bed_net_by_region)
```

- Pie plot
```{r echo=FALSE}
# your code

count <- table(selected_columns$WealthIndex)
pc<-round(count/sum(count)*100,2)
pie(count,
    labels = paste0(names(count)," (", pc,"%)",sep=""),
    main = "Distribution of Wealth Index", 
    xlab = "Wealth Index",
    col = c("green","blue","navy","purple","lightgreen")
    )
```

- Histogram
```{r}
malaria_by_district <- selected_columns %>%
  group_by(District, MalariaMeasurementResult) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  # Sort districts by total malaria cases
  group_by(District) %>%
  mutate(TotalCases = sum(Count)) %>%
  ungroup() %>%
  arrange(desc(TotalCases)) %>%
  # Select top 15 districts for readability
  slice_head(n = 15) %>%
  ggplot(aes(x = as.character(District), y = Count, fill = as.character(MalariaMeasurementResult))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Malaria Measurement Results in Top 15 Districts",
       x = "District", 
       y = "Number of Cases",
       fill = "Malaria Result") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(malaria_by_district)
```

- Boxplot
```{r}
children_bednet_boxplot <- selected_columns %>%
  mutate(ChildrenUnder5 = as.numeric(as.character(ChildrenUnder5))) %>%
  ggplot(aes(x = as.character(BedNetPresence), y = ChildrenUnder5, fill = as.character(BedNetPresence))) +
  geom_boxplot() +
  labs(title = "Number of Children Under 5 by Bed Net Presence",
       x = "Bed Net Presence", 
       y = "Number of Children Under 5") +
  theme_minimal()
children_bednet_boxplot
```

STATISTICAL ANALYSIS
```{r}
library(tidyverse)
library(stats)
```
1. Filtering out people who had Malaria measurement.
```{r}
# Function to filter malaria-related data
# Removes rows with missing malaria test results
filter_malaria_data <- function(data) {
  # Filter out rows with complete malaria test results
  filtered_data <- data %>%
    filter(!is.na(BloodSmearMalariaResult) & !is.na(RapidTestMalariaResult))
  
  # Print filtering summary
  cat("Data Filtering Summary:\n")
  cat("Total original records:", nrow(data), "\n")
  cat("Records after filtering:", nrow(filtered_data), "\n")
  
  return(filtered_data)
}
malaria_filtered_data <- filter_malaria_data(selected_columns)

```
2: Prevalence Calculation 
```{r}
# Function to calculate malaria prevalence for different test methods
calculate_malaria_prevalence <- function(data) {
  # Blood Smear Malaria Prevalence
  blood_smear_prevalence <- data %>%
    group_by(BloodSmearMalariaResult) %>%
    summarise(
      total_cases = n(),
      positive_cases = sum(BloodSmearMalariaResult == "Positive"),
      prevalence = (positive_cases / total_cases) * 100
    )
  
  # Rapid Test Malaria Prevalence
  rapid_test_prevalence <- data %>%
    group_by(RapidTestMalariaResult) %>%
    summarise(
      total_cases = n(),
      positive_cases = sum(RapidTestMalariaResult == "Positive"),
      prevalence = (positive_cases / total_cases) * 100
    )
  
  # Print prevalence results
  cat("\nBlood Smear Malaria Prevalence:\n")
  print(blood_smear_prevalence)
  
  cat("\nRapid Test Malaria Prevalence:\n")
  print(rapid_test_prevalence)
  
  return(list(
    blood_smear = blood_smear_prevalence, 
    rapid_test = rapid_test_prevalence
  ))
}

# Calculate prevalence
malaria_prevalence <- calculate_malaria_prevalence(malaria_filtered_data)
```

 3: District-Level Analysis 
 
```{r}
# Function to analyze malaria prevalence at district level
analyze_district_prevalence <- function(data) {
  district_prevalence <- data %>%
    group_by(District) %>%
    summarise(
      total_cases = n(),
      blood_smear_positive = sum(BloodSmearMalariaResult == "Positive"),
      rapid_test_positive = sum(RapidTestMalariaResult == "Positive"),
      blood_smear_prevalence = (blood_smear_positive / total_cases) * 100,
      rapid_test_prevalence = (rapid_test_positive / total_cases) * 100,
      average_altitude = mean(as.numeric(Altitude), na.rm = TRUE)
    ) %>%
    arrange(desc(blood_smear_prevalence))
  
  # Print district-level analysis
  cat("\nDistrict-Level Prevalence Analysis:\n")
  print(district_prevalence)
  
  return(district_prevalence)
}

# Perform district-level analysis
district_prevalence_results <- analyze_district_prevalence(malaria_filtered_data)
```

 4: Statistical Comparison 

```{r}
# Function to compare different test methods statistically
compare_test_methods <- function(data) {
  # Convert results to numeric for statistical testing
  blood_smear_numeric <- as.numeric(data$BloodSmearMalariaResult == "Positive")
  rapid_test_numeric <- as.numeric(data$RapidTestMalariaResult == "Positive")
  
  # Perform t-test
  test_comparison <- t.test(blood_smear_numeric, rapid_test_numeric)
  
  # Print test comparison results
  cat("\nTest Method Comparison (T-Test):\n")
  print(test_comparison)
  
  return(test_comparison)
}

# Compare test methods
test_method_results <- compare_test_methods(malaria_filtered_data)

##OBSERVATION:

# Statistical analysis reveals a highly significant difference between blood smear and rapid test results, 
# with the rapid test showing a slightly higher mean value and an extremely low p-value (5.579e-13).
```

 Determining the correlation between altitude and malaria prevalence

```{r}
analyze_altitude_correlation <- function(district_data) {
  # Linear regression model
  altitude_prevalence_model <- lm(
    blood_smear_prevalence ~ average_altitude, 
    data = district_data
  )
  
  # Visualization of Altitude vs Prevalence
  altitude_prevalence_plot <- ggplot(district_data, 
                                     aes(x = average_altitude, y = blood_smear_prevalence)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE) +
    labs(
      title = "Malaria Prevalence vs District Altitude",
      x = "Average District Altitude",
      y = "Blood Smear Malaria Prevalence (%)"
    ) +
    theme_minimal()
  
  # Print model summary
  cat("\nAltitude-Prevalence Relationship:\n")
  print(summary(altitude_prevalence_model))
  
  # Print the plot
  print(altitude_prevalence_plot)
  
  return(list(model = altitude_prevalence_model, plot = altitude_prevalence_plot))
}

# Analyze altitude correlation
altitude_correlation_results <- analyze_altitude_correlation(district_prevalence_results)

###OBSERVATION:

# The linear regression model suggests a negative relationship between average district altitude and malaria prevalence, but this relationship is not statistically significant (p-value = 0.2095).

# The scatterplot visualizes the negative trend, but the high degree of variability indicates that altitude alone is not a strong predictor of malaria prevalence.

# While the analysis hints at lower prevalence in higher altitude districts, the lack of statistical significance means the altitude-prevalence relationship cannot be conclusively determined from this data.

```






  







