---
title: "Assignement 2"
output:
  word_document: default
  pdf_document: default
date: "2024-01-10"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction :

This assignment consists of performing data cleaning and manipulation, and then some statistical analysis. 

## Dataset:

The dataset is retrieved from [Rwanda DHS (Demographic and Health Survey)](https://dhsprogram.com/Data/) 2020. The type of dataset used here is Household member. You will get data in two files: main SPSS File and Map File (for descriptions). 

Your Assignments steps:

1. Read the dataset in R.

```{r echo=FALSE}
# required package
library(haven)

# import data
data_sav <- read_sav(" ")
#data_map <- read.table("DHS2020/RWPR81FL.MAP", sep = "\t", header = TRUE)

# data dimension
dim_no <- dim(data_sav)
```
## In our dataset of hoursehourd We have `r dim_no[1]` total number of Obersvation and `r dim_no[2]` total number of Variables.


- Visualize, inspect and get familiar with the data
```{r echo=FALSE}
# your code here
View(data_sav)
 #View(data_map)

```


2. Select only few columns, important in this Assignments. They are the following: "HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7",
  "HML10", "HML22", "HML32","HML33", "HML35"

```{r echo=FALSE}

# Select columns by their names
selected_columns <- data_sav[, c("HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7","HML10", "HML22", "HML32","HML33", "HML35")]
selected_columns

```

  
3. Rename variables using the variable descriptions below. Give meaningful (short) name to the variables of your choice.

>* HV001= "Cluster number", 
>* HV009 = "Number of household members",
>* HV010 = "Number of eligible women in household",
>* HV011 = "Number of eligible men in household",
>* HV014 = "Number of children 5 and under (de jure)",
>* SHDISTRICT = "District (geographic area)",
>* HV024 = "Region (provinces, corresponding values in a map file)",
>* HV025 = "Type of place of residence (rural versus urban)",
>* HV040 = "Cluster altitude in meters",
>* HV227 = "Presence of mosquito bed net for sleeping",
>* HV228 = "Number of children under 5 who slept under a mosquito bed net",
>* HV270 = "Wealth index combined (an index based on various household assets indicating socio-economic status)",
>* HV105 = "Age of household members",
>* HV106 = "Highest educational level attained by individuals",
>* HML3 = "Net observed by interviewer",
>* HML4 = "Months ago the net was obtained",
>* HML7 = "Brand of net",
>* HML10 = "Insecticide-Treated Net (ITN)",
>* HML22 = "Obtained net from campaign, antenatal, or immunization visit",
>* HML33= "Result of malaria measurement", 
>* HML32 = "Final result of malaria from blood smear test",
>* HML35 = "Result of malaria rapid test"


# Renaming the dataset variables with meaningful names
```{r echo=FALSE}
#Rename variables using the variable descriptions given.
colnames(selected_columns) <- c(
  "ClusterNumber", "HouseholdMembers", "EligibleWomen", "EligibleMen", "ChildrenUnder5", 
  "District", "Region", "ResidenceType", "Altitude", "BedNetPresence", 
  "ChildrenUnder5BedNet", "WealthIndex", "Age", "EducationLevel", 
  "NetObserved", "NetObtainedMonthsAgo", "NetBrand", "ITN", "NetSource", 
  "BloodSmearMalariaResult", "MalariaMeasurementResult", "RapidTestMalariaResult"
)
head(selected_columns)

```


## Data cleaning

1. Inspect each variables, decode variable to its original unique variables. 
Example, Variable "HV024"(Region) has Unique values 1,2,3,4,5. Decode it to orginal Region Kigali, South, West, North, East
Use Map file to see the description of each values in data. 

```{r echo=FALSE}
# your code

# Decode all columns
all_columns <-  c(
  "ClusterNumber", "HouseholdMembers", "EligibleWomen", "EligibleMen", "ChildrenUnder5", "District", "Region", "ResidenceType", "Altitude", "BedNetPresence","ChildrenUnder5BedNet", "WealthIndex", "Age", "EducationLevel", 
  "NetObserved", "NetObtainedMonthsAgo", "NetBrand", "ITN", "NetSource", 
  "MalariaMeasurementResult", "BloodSmearMalariaResult", "RapidTestMalariaResult")

selected_columns[all_columns] <- sapply(selected_columns[all_columns], as_factor)


# region_name_decord <- as.character(selected_columns$Region)
# for (m in 1:length(region_name_decord)) {
#   
#   if (region_name_decord[m] == "1"){
#     region_name_decord[m] <- "Kigali"
#   }else if (region_name_decord[m] == "2") {
#     region_name_decord[m] <- "South"
#   }else if (region_name_decord[m] == "3") {
#     region_name_decord[m] <- "West"
#   }else if (region_name_decord[m] == "4") {
#     region_name_decord[m] <- "North"
#   }else {
#     region_name_decord[m] <- "East"
#   
#   }
# }


```


<!-- ```{r echo=FALSE} -->
<!-- selected_columns$ResidenceType -->

<!-- residence_type <- as.numeric(selected_columns$ResidenceType) -->
<!-- for (i in 1:length(residence_type)) { -->
<!--   if(residence_type[i]=="1"){ -->
<!--    residence_type[i]<-"Urban" -->
<!--   }else{ -->
<!--     residence_type[i]<-"Rural"  -->
<!--   } -->
<!-- } -->
<!-- selected_columns$ResidenceType -->
<!-- ``` -->


2. Handling Missing Values:

Determine columns with missing values. Devise the strategy to handle missing values: Deleting missing values, replacing missing values with mean or mode.
```{r echo=FALSE}
# your code
# checking columns with missing values in entire datasets

missing_value <- colSums(is.na(selected_columns))

# check columns with missing values
missing_value[missing_value > 0]

```
```{r echo=FALSE}
# cleaning missing values using Mode

mode <- function(x) names(sort(table(x), decreasing = TRUE))[1]
selected_columns$ChildrenUnder5BedNet[is.na(selected_columns$ChildrenUnder5BedNet)] <- mode(selected_columns$ChildrenUnder5BedNet)
head(selected_columns)
```
```{r echo=FALSE}
# # Define the mode function
# mode <- function(y) names(sort(table(y), decreasing = TRUE))[1]
# 
# # Check if there are any NA values in the column before replacing
# if (any(is.na(selected_columns$NetObserved))) {
#   selected_columns$NetObserved[is.na(selected_columns$NetObserved)] <- mode(selected_columns$NetObserved)
# }
# 
# # View the first few rows of the updated dataframe
# head(selected_columns)


# Mode function to handle categorical data
calculate_mode <- function(x) {
  # Handle NA values
  x <- x[!is.na(x)]
  # Find the most frequent value
  if (length(x) == 0) return(NA)  # If all values are NA, return NA
  mode_value <- names(sort(table(x), decreasing = TRUE))[1]
  return(mode_value)
}
# Handle all using lappy function
selected_columns[] <- lapply(selected_columns, function(col) {
  if (is.numeric(col)) {
    # Replace numerical missing values with the mean
    col[is.na(col)] <- mean(col, na.rm = TRUE)
  } else if (is.factor(col) || is.character(col)) {
    # Replace categorical missing values with the mode
    col[is.na(col)] <- calculate_mode(col)
  }
  return(col)
})

View(selected_columns)
```


3. Create new variables

>a. Create variable called "Old Mosquito" variable HML4 (Months ago the net was obtained). The created variable must binary with 1 when mosquito is more than 24 months old. 

```{r echo=FALSE}
# Create the "Old Mosquito" variable
selected_columns<- selected_columns %>%
  mutate(Old_Mosquito = ifelse(NetObtainedMonthsAgo > 24, 1, 0))

# View the updated data
selected_columns
```


>b. Create Variable "Average District altitude". Create this variable by averaging cluster altitude in each district. We have three variables HV001= "Cluster number", SHDISTRICT = "District (geographic area)" and HV040 = "Cluster altitude in meters". Filter out clusters in each district, do `mean` of cluster
altitude in that district. 

```{r echo=FALSE}
# Load necessary libraries
library(dplyr)

# Assume your data is in a data frame called 'data'
# Example data
# data <- data.frame(HV001 = c(1,2,3,4), SHDISTRICT = c("A", "A", "B", "B"), HV040 = c(500, 600, 700, 800))

# Create the Average District Altitude variable

selected_columns<- selected_columns %>%
  group_by(District) %>%
  mutate(Average_District_Altitude = mean(as.numeric(Altitude))) %>% ungroup()

# View the data
print(selected_columns)

```


## Data visualizations: 

Produce visualization of your choice. At least each of these
- Bar plot

```{r echo=FALSE}
# Count the number of malaria cases by district
malaria_count <- table(selected_columns$District, selected_columns$MalariaMeasurementResult)

# For each district, count how many malaria measurement results (1) are present
# Assuming 'MalariaMeasurementResult == 1' means a positive malaria result
malaria_results_by_district <- apply(malaria_count, 1, function(x) x["1"])

#Hundle NA
malaria_results_by_district[is.na(malaria_results_by_district)] <- 1
# Create a bar plot
barplot(malaria_results_by_district, 
        main = "Malaria Results by District", 
        col = "blue", 
        ylab = "Number of Malaria Results", 
        xlab = "District")

# Add the counts as labels on top of the bars
text(x = seq_along(malaria_results_by_district),
     y = malaria_results_by_district,
     labels = malaria_results_by_district,
     pos = 3, cex = 1, col = "black")
```

```{r}
# your code
count <- table(selected_columns$District)
barplot(count,
        main = "Distribution of District", 
        col = "blue")

```

- Pie plot
```{r echo=FALSE}
# your code

count <- table(selected_columns$WealthIndex)
pc<-round(count/sum(count)*100,2)
pie(count,
    labels = paste0(names(count)," (", pc,"%)",sep=""),
    main = "Distribution Wealth Index", 
    xlab = "Wealth Index",
    col = c("green","blue","navy","purple","lightgreen")
    )
```

- Histogram
```{r echo=FALSE}
# # your code
# count_att <- table(selected_columns$Average_District_Altitude)
# hist(count_att,
#      main = "Distribution Of Average District Altitude", 
#         col = "lightblue", 
#         xlab = "District"
#      )

malaria_by_district <- selected_columns %>%
  group_by(District, MalariaMeasurementResult) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  # Sort districts by total malaria cases
  group_by(District) %>%
  mutate(TotalCases = sum(Count)) %>%
  ungroup() %>%
  arrange(desc(TotalCases)) %>%
  # Select top 15 districts for readability
  slice_head(n = 15) %>%
  ggplot(aes(x = as.character(District), y = Count, fill = as.character(MalariaMeasurementResult))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Malaria Measurement Results in Top 15 Districts",
       x = "District", 
       y = "Number of Cases",
       fill = "Malaria Result") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(malaria_by_district)
```

- Boxplot
```{r}
# your code

```


## Statistical analysis

### Descriptive statistics

1. Use Variable "HML33" to filter out people who had Malaria measurement. 
```{r}
# your code
colnames(selected_columns)
table(selected_columns$MalariaMeasurementResult== "Measured")

```

2. Calculate Malaria [Prevalence](https://dhsprogram.com/data/Guide-to-DHS-Statistics/Prevalence_of_Malaria_in_Children.htm) for both "Blood Smear" and "Rapid Test"
```{r echo=FALSE}
# Calculate prevalence for Blood Smear
total_blood_smeared <- sum(!is.na(selected_columns$BloodSmearMalariaResult))  # Total number of tests (excluding NA)

positive_blood_smear <- sum(selected_columns$BloodSmearMalariaResult == "Positive")  # Number of positive results

prevalence_blood_smeared <- (positive_blood_smear / total_blood_smeared) * 100

# Calculate prevalence for Rapid Test
total_rapid_tested <- sum(!is.na(selected_columns$RapidTestMalariaResult))  # Total number of rapid tests (excluding NA)
positive_rapid_test <- sum(selected_columns$RapidTestMalariaResult == "Positive")  # Number of positive results

prevalence_rapid_test <- (positive_rapid_test / total_rapid_tested) * 100

# Print results
cat("Malaria Prevalence for Blood Smear: ", prevalence_blood_smeared, "%\n")
cat("Malaria Prevalence for Rapid Test: ", prevalence_rapid_test, "%\n")
```

3. Aggregate Prevalence at district Level
```{r echo=FALSE}
# Prevalence by district (example: grouping by district)
prevalence_by_district <- selected_columns %>%
  group_by(District) %>%
  summarise(
    prevalence_blood_smeared = (table(BloodSmearMalariaResult == "Positive") / sum(!is.na(BloodSmearMalariaResult))) * 100,
    prevalence_rapid_test = (sum(RapidTestMalariaResult == "Positive") / sum(!is.na(RapidTestMalariaResult))) * 100
  )

# Print results
print(prevalence_by_district)
```

### Analytical Analysis
1. Compare the prevalence in both tests and state if they are different.

***Hint:*** Check `?` the documentations for `t.test` and `aov`.
```{r}
# your code
```

### Bonus

2. Using a statistical model of your choice, determine if there is a relationship between malaria prevalence in a district and its average altitude. 
```{r}
# your code
```