---
title: "Prediction_Model"
author: "Josue Iradukunda"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
x <- 3
ifelse(x < 4, T, F)  # this is a vectorized function (applies on vectors)
# Alternative 1
if (x < 4) print(TRUE) else print(FALSE)


if (x < 4){
  print(TRUE)
}else {
  print(FALSE)
}
```
We can embed `if` to `if` and `else`
```{r echo=FALSE}
set.seed(123)
(marks <- round(rnorm(50, mean = 60, sd = 15), 2))
```
Exercise: Write a if statement to check for each value of marks if they fail in the following categories.
*0 -49 -> fail
* > 49 - 59 ->pass
* > 69 - 79 -> Good pass
* > 79 - 84 -> Very good pass
* > 84 -> Distinction

1. Taking Values of `marks` Direct

```{r echo=FALSE}
category <- character()
set.seed(123)
(marks <- round(rnorm(50, mean = 60, sd = 15), 2))

for (m in marks) {
  print(m)
if (m >= 0 && m< 60) {
  print("Fail")
  category <- c(category, "Fail")
}else if ((m >= 60 && m < 70) ) {
  print("Good Pass")
  category <- c(category, "Good Pass")
} else if (m >= 70 && m < 80) {
  print("Very Good Pass")
  category <- c(category, "Very Good Pass")
}else {
  print("Distinction")
  category <- c(category, "Distinction")
}
}

```
Using Indexes

```{r echo=FALSE}
category2 <- character()
for (i in 1:length(marks)) {
  print(marks[i])
  if (marks[i] < 60){
    print("Fail")
    category2 [i] <- "F"
  }else if (marks[i] >= 60 && marks[i] > 70) {
    print("Good Pass")
   category2 [i] <- "GP"
  } else  if (marks[i] >= 70 && marks[i] < 80) {
    print("Very Good Pass")
   category2 [i] <- "VGP"
  }else{
    print("Distinction")
    category2 [i] <- "D"
  }
  
}
head(df_marks <- data.frame(marks, category2),10)
```
```{r echo=FALSE}
head(df_marks <- data.frame(marks, category),50)

```

3. Using `ifelse` (Vectorized function)

```{r}
category3 <- ifelse(marks >= 0 & marks < 60, "Fail",
                    ifelse (marks >= 60 & marks < 70, "Pass",
                            ifelse(marks >= 70 & marks < 80, "Good Pass", 
                                   ifelse(marks >= 80 & marks < 85, "Very Good Pass", "Distinction"))))

print(category3)
```

# **Assignment:** Write an R function that will take as an input a vector or marks and `return` a data frame of marks and categories (`F`, `P`, `GP`,`VGP`, and `D`). The function should be able to make some plots (`pie`, `barplot`)

```{r echo=FALSE}


# Define the function
classify_marks <- function(marks) {  
  # Initialize an empty character vector to store categories
  category2 <- character(length(marks))
  
  # Classify each mark
  for (i in 1:length(marks)) {
    if (marks[i] < 60) {
      category2[i] <- "F"   # Fail
    } else if (marks[i] >= 60 && marks[i] < 70) {
      category2[i] <- "GP"  # Good Pass
    } else if (marks[i] >= 70 && marks[i] < 80) {
      category2[i] <- "VGP" # Very Good Pass
    } else {
      category2[i] <- "D"   # Distinction
    }
  }
  # Create a data frame with marks and their corresponding categories
  df_marks <- data.frame(marks = marks, category2 = category2)
  
  # Return the data frame
  return(df_marks)
}
# Calling the function created
classify_marks(marks)

```

```{r echo=FALSE}
(cat_count <- table(df_marks$category))
 barplot(cat_count, main = "Category Distribution", col = "lightblue", xlab = "Category", ylab = "Frequency")
  
  # Create a pie chart for category distribution
  pie(cat_count, main = "Category Distribution (Pie Chart)", col = rainbow(length(cat_count)))

```


# function ask user to input data.


```{r}
# Initialize an empty vector to store the marks
marks <- c()
disply_mks <- function(){
# Ask the user for the number of marks they want to enter
num_marks <- as.integer(readline(prompt = "How many marks do you want to record?: "))

# Loop to collect the marks
for (i in 1:num_marks) {
  # Read mark from user
  mark <- as.integer(readline(prompt = paste("Enter mark", i, ": ")))

  # Add the mark to the vector
  marks <- c(marks, mark)
  
  # Categorize the mark using if-else statements
  if (mark < 60) {
    category <- "Fail"
  } else if (mark >= 60 && mark < 70) {
    category <- "Pass"
  } else if (mark >= 70 && mark < 80) {
    category <- "Good pass"
  } else if (mark >= 80 && mark < 85) {
    category <- "Very Good Pass"
  } else {
    category <- "Excellent"
  }

  
}
 # Print the mark and its category
  cat("Mark:", mark, "- Category:", category, "\n")
  # Optionally, print all marks entered
cat("All marks entered:", marks, "\n")

df_marks <- data.frame(marks = marks, category)
  
  # Return the data frame
return(df_marks)
}
disply_mks()


```

```{r echo=FALSE}
# Create a data frame with marks and their corresponding categories
  df_marks <- data.frame(marks = marks, category2 = category2)
  
  # Return the data frame
  return(df_marks)

```


