---
title: "Outliers_Topic"
author: "Josue Iradukunda"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
x <- c(19.103, 19.632, 22.494, 20.113, 10.5, 22.744, 20.737, 17.976, 
       18.901, 192.87, 21.959, 0.001, 20.641, 20.177, 19.111, 22.859, 
       207.97, 16.853, 21.122, 19.244, 18.291, 19.651, 18.358, 18.834, 19)

boxplot(x_copy <- x)
```


```{r echo=F}
Q <- as.numeric(quantile(x, probs = c(0.25, 0.5, 0.75)))
d <- as.data.frame(t(Q)) #t is for transpose
names(d) <- paste0("Q", 1:length(Q))
rownames(d) <- "stats"
d$IQR <- d$Q3 - d$Q1
d$LowerLim = d$Q1 - 1.5*d$IQR
d$UpperLim = d$Q3 + 1.5*d$IQR
d
```
Checking for all elements of x if they are outliers or not:

```{r echo=FALSE}
(d_out <- data.frame(x = x, outlier = x < d$LowerLim | x > d$UpperLim))

```

Identifying outliers:

-using the `d_out` table

```{r echo=FALSE}
x[d_out$outlier]
```
**using the boxplot() function

```{r echo=FALSE}
(bxpt <- boxplot(x, range = 1.5, plot = FALSE)) # we do not plot
```
The object bxpt is a list of 6 elements with a column named out which stands for outliers. You all know how to identify and element in a list using $ or [["the name of the element"]].

```{r echo=FALSE}
x[d_out$outlier] # ours
bxpt$out
bxpt[["out"]]
```
Now, we can consider all the outliers as missing values (NA). First, we need to locate them.

*Method 1
```{r echo=FALSE}
# find all outliers
x[d_out$outlier] <- NA # or x[x %in% bxpt$out]
(x[d_out$outlier] <- mean(x, na.rm = TRUE))
```
*Method 2

```{r eval=FALSE}
x[d_out$outlier] <- mean(x[d_out$outlier])
```

```{r}
# boxplot of imputed x
par(mfrow = c(1, 2))
boxplot(x_copy, main = "Before\nhandling outliers", cex.main = 0.7)
boxplot(x, main = "After\nhandling outliers", col = "green", cex.main = 0.7)
```

# Data Simultion and Visualization
## Charts
## Bar charts/plot is used in counts

# horiz = TRUE it plot in horizontal way

```{r echo=FALSE}
frequencies <- c(Female = 35, Children = 25, Male = 40)
barplot(frequencies, names.arg = names(frequencies), cex.names=0.7, 
        horiz = TRUE, col = as.factor(names(frequencies))) 
#col = as.factor(names(frequencies)))  different colors

```

## Pie chart/plot

```{r echo=FALSE}
par(mfrow = c(1, 2))
pie(
  frequencies, cex = 0.5, labels = paste0(names(frequencies), "(", frequencies, ")"), col = c("purple","yellow", "blue")
  
)

pie(frequencies, cex = 0.8, labels = paste0(names(frequencies), "(", frequencies, ")"), col = rainbow(length(frequencies)))
```
## 4.1.3 Histograms

```{r echo=FALSE}
set.seed(12092024)
par(mfrow = c(2, 1))
x <- sample(1:120, size = 1000, replace = TRUE); 
hist(x, probability = TRUE) # use probability = TRUE to have densities

       # instead of counts (frequencies)
# Density plots
plot(density(x))
```

```{r}
set.seed(12092024)
x <- sample(1:120, size = 1000, replace = T);
par(mfrow = c(1, 2))
hist(x, probability = T) # use probability = TRUE to have densities
plot(density(x)) # density plot
```

## Overlapping with density
```{r}
hist(x, probability = T, col=colors()[10])
lines(density(x), col="blue",lwd=3, lty=4)
```

## Scatter plot

```{r echo=FALSE}
x <- seq(0, 2*pi, le = 50)
y <- sin(x)
#z <- cos(x)
#tg <- tan(x)
plot(x, y, pch = 16, col = "blue")
```
Exercise:
Make a scatter plot of `petal.Length` and `petal.Width` colored by `Species` using `iris` data frame

```{r}
names(iris)
```

