---
title: "Dimension"
author: "Josue Iradukunda"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dimension: row position, column position, and matrix level

### Lists

A list is a collection of objects of defferent types. The sizes of elements could be different.

```{r}
A <- matrix(c(2, 1, 0, 3), ncol = 2, byrow = TRUE)
mylist <- list("matrix" = A, "sequence" = 5, "Bool" = TRUE)

# mylist$matrix
```
```{r}
L <- list()
A <- matrix(0,5,6)
L$A <- A #add on list a values.
print(L)

#Adding sequence on list
L$my_seq <- sample(10, 5) #second number should be lest the firs, if not its return error.
print(L)

# add a boolean

L$Bool <- TRUE
print(L)

L[['iris']] <- head(iris, 10)
print(L) # print 10 element of list

# checking number of elements in R

length(L)

L[[5]] <- "I am learning R programming"
print(L)
```

Getting the `names` of element
```{r}
names(L)

```
Renaming the 5th element of L
```{r}
names(L)[5]  <- "String"
print(L)
```
# Delete element in a List 
```{r}
L$Bool <-NULL
print(L)
```
# Data Frame data type

Is a table of any number of $n$ number of rows (observations) and $n$ number of columns (features or or variables) Variables can take any data type.

### Converting a continous variable into a categorical variable

```{r echo=TRUE}
set.seed(123)
n <- 120

dt <- data.frame(
  "ID" = paste("Participant", 1:n),
  "age" = sample(0:120, size = n)
)
head(dt)

```
```{r}
(brks <- seq(0,120, by = 10))

dt$age_group <- cut(dt$age, breaks = brks, include.lowest = TRUE)

head(dt, 8)
```

# Exercise
1. provide the counts by age group

```{r}
count_age_group <- table((dt$age_group))
count_age_group
```
# make a barplot of age group
```{r}
barplot(count_age_group)
```
#make histogram of ages 

```{r}
hist(age, col = "blue")

```
#Display the ID's of participants in the $70,80$ age group

```{r}
Id_group <- dt[dt$age_group == "(70,80]", "ID" ]
Id_group
```
```{r}
l <- split(dt, dt$age_group)
l$`(70,80]`$ID  
```

